Analyzing Student Use of Spacing and Interleaving Strategies in Interactions with GenAI-Powered Chatbots in Programming Courses

### 2024-11-14 CSEDU2025/Paper.tex
update paper

This section presents the theoretical framework that underpins the study. First,
we introduce the concept of metacognition used in this study. Next, we discuss
related work exploring the use of metacognitive strategies and scaffolds
in educational settings.

@@

\subsection{Related Work}

@@

\cite{Bastani24} investigated the impact of generative artificial intelligence
(AI), specifically GPT-4, on human learning, with a focus on mathematics
education in a high school. The problem addressed was how the use of generative
AI could affect the acquisition of new skills, which is crucial for long-term
productivity. The method involved a controlled experiment with approximately
one thousand students, who were exposed to two GPT-4-based tutors: a simple
tutor (GPT Base) and another with safeguards designed to promote learning
(GPT Tutor). The results showed that while access to GPT-4 improved performance
on practice exercises (48\% with GPT Base and 127\% with GPT Tutor), the removal
of access to GPT Base led to a 17\% decrease in student performance on exams.
This indicated that the unrestricted use of *GPT Base* could hinder learning.
However, the GPT Tutor was able to mitigate this negative effect.

\cite{Boudouaia24} investigated the impact of ChatGPT-facilitated programming on
college students’ programming behaviors, performance, and perceptions. The
problem addressed was whether the use of ChatGPT in programming education could
influence how students interacted with programming tasks, their learning
outcomes, and their attitudes towards technology. The study employed a
quasi-experimental design involving 82 college students divided into two groups:
one using ChatGPT-assisted programming (CFP) and the other using traditional
self-directed programming (SDP). The method combined mixed data collection
techniques, including behavioral logs, programming performance evaluations, and
interviews. The results showed that students using the CFP mode exhibited more
frequent behaviors such as debugging and reading feedback. Although CFP students
had a higher average score in programming performance, there was no
statistically significant difference compared to the SDP group. However,
students’ perceptions of ChatGPT improved significantly after the intervention,
with increased perceived usefulness, ease of use, and intention to continue
using the tool in the future.

@@

The analysis phase involved the use of Python scripts to extract information
from previously classified data. The datasets and code used in this process are
available at: https://github.com/rodrigoprestesmachado/uc3m.

@@

Student corrections to chatbot responses accounted for 4.2\% of the interactions.
In most cases, these corrections occurred when the student already had the
correct answer to an exercise and identified an error in the chatbot response,
as illustrated in conversation 2 in Figure \ref{fig:graph2}. These corrections
highlight that, although the chatbot provides quick feedback, it is not always
accurate.

@@

Despite some occasional variation (3.54), prompts classified as Uncategorized
exhibit a relatively low average time (1.65) for students to resume interaction
with the chatbot. This variation occurs, in part, due to situations where students
end the conversation with an uncategorized message, such as a simple
\textit{"thanks"} but later choose to resume the interaction.

When a student requests the chatbot to change the language (LC) in a response, the
new interaction happens quickly (1.89) with little variation in response time
(3.28), meaning it is predictable that the student will continue interacting.

@@

Conceptual questions (CQ) generally result in a high average time (5.41) for
students to re-engage with the chatbot. Moreover, the high standard deviation
(10.54) indicates an unpredictable variation in response times. This
unpredictability may be attributed to the fact that, in some cases, the
complexity of the question and answer demands more reflection time from the
student, whereas in others, lesser complexity allows for a quicker interaction
with the chatbot.

Finally, a message classified as a Complete Solution request (SCO) exhibits the
longest average time (6.55) for students to resume interaction with the chatbot.
Additionally, the high standard deviation (10.14) highlights significant
unpredictability in response time. After requesting a complete solution,
students tend to take longer to return, due to the complexity of the problems
involved, such as the implementation of sorting algorithms like Heapsort. Table
\ref{tab:averages} presents the means and standard deviations for each
category.

@@

The theory of metacognition, along with the classification of student actions,
proved effective in understanding and revealing how students utilize the chatbot.
The results showed that most messages fall into the categories of Conceptual
Questions and Debugging Help, both known for promoting slower engagement and
potentially fostering more active learning. However, the results also indicated
that students' use of study strategies such as spacing and interleaving remains
limited, from which we can conclude that encouraging students to apply these
strategies more intentionally could enhance their learning while using GenAI.

@@

To further explore this, a broader study is planned to examine, for instance,
the relationship between student profiles, use of metacognitive scaffolds, their
effective study strategies in using a GenAI and, learning outcomes.
Additionally, to support educators in analyzing student behavior and providing
more targeted assistance, an AI model is being developed to automatically
classify student messages, offering valuable insights into how these tools are
being used.

>>>>

This section outlines the theoretical framework that grounds the study. We
begin by introducing the concept of metacognition, along with the spacing and
interweaving strategies applied in this research. Next, we examine related works
 organized into two subsections: the use of GenAI in programming and recent
 studies emphasizing the importance of metacognition in the use of GenAI.

@@

\subsection{GenAI and Programming}

@@

\cite{Bastani24} investigated the impact of GenAI, specifically GPT-4, on human
learning, with a focus on mathematics education in a high school. The problem
addressed was how the use of generative AI could affect the acquisition of new
skills, which is crucial for long-term productivity. The method involved a
controlled experiment with approximately one thousand students, who were exposed
to two GPT-4-based tutors: a simple tutor (GPT Base) and another with safeguards
designed to promote learning (GPT Tutor). The results showed that while access
to GPT-4 improved performance on practice exercises (48\% with GPT Base and
127\% with GPT Tutor), the removal of access to GPT Base led to a 17\% decrease
in student performance on exams. This suggests that unrestricted use of
\textit{GPT Base} could hinder learning. However, the GPT Tutor was able to
mitigate this negative effect.

\cite{Boudouaia24} investigated the impact of ChatGPT-facilitated programming on
college students’ programming behaviors, performance, and perceptions. The
problem addressed was whether the use of ChatGPT in programming education could
influence how students interacted with programming tasks, their learning
outcomes, and their attitudes towards technology. The study employed a
quasi-experimental design involving 82 college students divided into two groups:
one using ChatGPT-assisted Programming (CAP) and the other using traditional
self-directed Programming (SDP). The method combined mixed data collection
techniques, including behavioral logs, programming performance evaluations, and
interviews. The results showed that students using the CAP mode exhibited more
frequent behaviors such as debugging and reading feedback. Although CAP students
had a higher average score in programming performance, there was no
statistically significant difference compared to the SDP group. However,
students’ perceptions of ChatGPT improved significantly after the intervention,
with increased perceived usefulness, ease of use, and intention to continue
using the tool in the future.

\subsection{GenAI and Metacognition}

@@

The analysis phase involved the use of Python scripts to extract information
from previously classified data. The data sets and code used in this process are
available at: https://github.com/rodrigoprestesmachado/uc3m.

@@

Student corrections to chatbot responses represented 4.2\% of the interactions.
In most cases, these corrections occurred when the student already had the
correct answer to an exercise and identified an error in the chatbot response,
as illustrated in conversation 2 in Figure \ref{fig:graph2}. These corrections
highlight that, although the chatbot provides quick feedback, it is not always
accurate.

@@

Despite some occasional variation (3.54), prompts classified as Uncategorized
exhibit a relatively low average time (1.65) for students to resume interaction
with the chatbot. This variation occurs, in part, because of situations where students
end the conversation with an uncategorized message, such as a simple
\textit{"thanks"}, but later choose to resume the interaction.

When a student requests the chatbot to change the language (LC) in a response, the
new interaction occurs quickly (1.89) with little variation in response time
(3.28), meaning that it is predictable that the student will continue interacting.

@@

Conceptual questions (CQ) generally result in a high average time (5.41) for
students to reengage with the chatbot. Moreover, the high standard deviation
(10.54) indicates an unpredictable variation in response times. This
unpredictability may be attributed to the fact that, in some cases, the
complexity of the question and answer demands more reflection time from the
student, whereas in others, lesser complexity allows for a quicker interaction
with the chatbot.

Finally, a message classified as a Complete Solution request (SCO) exhibits the
longest average time (6.55) for students to resume interaction with the chatbot.
Additionally, the high standard deviation (10.14) highlights significant
unpredictability in response time. After requesting a complete solution,
students tend to take longer to return due to the complexity of the problems
involved, such as the implementation of sorting algorithms such as Heapsort. Table
\ref{tab:averages} presents the means and standard deviations for each
category.

@@

The metacognition theory, along with the classification of student actions,
proved to be effective in understanding and revealing how students use the chatbot.
The results showed that most messages fall into the categories of Conceptual
Questions and Debugging Help, both known for promoting slower engagement and
potentially fostering more active learning. However, the results also indicated
that students' use of study strategies such as spacing and interleaving remains
limited, from which we can conclude that encouraging students to apply these
strategies more intentionally could enhance their learning while using GenAI.

@@

To further explore this, a broader study is planned to examine, for instance,
the relationship between student profiles, use of metacognitive scaffolds, their
effective study strategies in using a GenAI, and learning outcomes.
In addition, to help educators analyze student behavior and provide
more targeted assistance, an AI model is being developed to automatically
classify student messages, offering valuable information on how these tools are used.
